Estimation of Intracortical Connectivity in a Dynamical Neural Field Model

### 2010-04-13 ltx/BrainIDE.tex
changes to coloured noise

where $T$ is the time step, $\lambda = 1-\alpha T$ and $e_t(\mathbf{r}) \sim \mathcal{GP}\left(0,\Sigma\left(\mathbf{r}-\mathbf{r}'\right)\right)$ 
is an iid spatially colored, temporally white disturbance term. This term was added to drive the system and account for uncertainty in the model. To generate iEEG/LFP data we used the output function 

@@

where $\mathbf{r}_n$ defines the location of the sensors in the field where $n=1,...,N$ indexes the sensors and $\varepsilon_t \sim \mathcal{N}\left(0,\Sigma_y\right)$. The output kernel $m(\mathbf{r}-\mathbf{r}')$ governs the geometry of pick-up range of sensors where

@@

where $\mathbf{e}_t$ is a zero-mean normal white noise term with the covariance (see Appendix D)

@@

\boldsymbol{\Sigma}_w=\boldsymbol{\Gamma}^{-1}\iint\limits_{\Omega}\boldsymbol{\phi}\left(\mathbf r\right) \rho\left(\mathbf{r}- \mathbf{r}' \right)\boldsymbol{\phi}\left(\mathbf{r}'\right)^{\top}d\mathbf{r}' d\mathbf{r}\boldsymbol{\Gamma}^{- \top} 
\rho\left(\mathbf{r}-\mathbf{r'}\right), & \tau=0 \\
\label{eq:FieldCovariance}

@@

The approximation of the state density is made by propagating the so called sigma points through the state-space model. The sigma points describe the mean and covariance of the state estimate. The accuracy of the approximation is linked to the time step used in the temporal discretization of the system, where the state transition appears more linear with a finer resolution. To explain the operation of the UKF we begin by deriving the traditional Kalman filter with the state-space model
\begin{eqnarray}
	\mathbf{x}_t = q(\mathbf{x}_t)\boldsymbol{\theta} - \lambda\mathbf{x}_t + k\left(u_t\left(\mathbf{r}\right)\right) + \mathbf{e}_t \\
	\mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t,
\end{eqnarray}
where $\boldsymbol{\theta}$ and $\lambda$ are the systems parameters. In forwards iteration since the observation equation is linear in parameters we only need to use Unsented transformation in prediction step. The update step equations remains the same as the standard Kalman filtering formulations. In backward smoothing path, estimation from the forward filtering is used to form the new set of sigma points in a similar way to the filterin step, however the predicted cross-covariance of the states need to be calculated. A summary of the algorithm is given in Algorithm

\section{Comparison of Estimated Parameters and True Parameters}

\begin{algorithm}
\caption{The Unscented Kalman Filter}\label{CPTEMDAlgorithm}
\begin{algorithmic}[1]
\State $k=1$
\State use CPT from Algorithm~\ref{CPTAlgorithm} to estimate $\phi_k(n)$ from $y(n)$
\State find local extrema of $cos\left(\phi_k(n)\right)$
\State use extrema to form the splines $s_{min,k}(n)$ and $s_{max,k}(n)$
\State calculate $x_{0,k}(n) = 0.5\left(s_{min,k}(n) + s_{max,k}(n)\right)$
\State find local maxima of $h_k(n)$
\State use maxima to form the spline $r_k(n) = s_{max,k}(n)$
\While { $x_{0,k}(n)$ is not monotonic }
    \State $k=k+1$
    \State perform CPT on $x_{0,k-1}(n)$ to estimate $\phi_{k}(n)$
    \State repeats steps 3 to 7 to approximate parameters
\EndWhile

\end{algorithmic}
\end{algorithm}

@@

\section{Discrete Time Model}\label{Time Discretization}
To form the IDE neural field model a time discretization must be perform. We used a one-step Euler method where Eq.~\ref{FinalForm1} can be approximated by
\begin{equation}\label{Euler Approximation}	
\frac{v\left( \mathbf{r},t+T \right) - v\left( \mathbf{r},t\right)}{T} =   -\alpha v\left( \mathbf{r},t \right) + \int_\Omega  {w\left( \mathbf{r}-\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)d\mathbf{r}'} + u\left(\mathbf{r},t\right).
\end{equation}
For clarity, we shall index time points in the discrete time form of the model using the subscript $t$ and the next time point as $t+1$. Rearranging Eq.~\ref{Euler Approximation} we get
\begin{eqnarray}\label{Euler Approximation}	
v_{t+1}\left( \mathbf{r}\right) &=& v_t\left( \mathbf{r}\right) -T \alpha v_t\left( \mathbf{r}\right) + T \int_\Omega  {w\left( \mathbf{r}-\mathbf{r}' \right)f\left( {v_t\left( \mathbf{r}'\right)} \right)d\mathbf{r}'} + T u_t\left(\mathbf{r}\right).
\end{eqnarray}
The discrete time form of the model is
\begin{equation}\label{Discrete Time Model1}
	v_{t+1}\left(\mathbf{r}\right) = \lambda v_t\left(\mathbf{r}\right) + T \int_\Omega { w\left(\mathbf{r}-\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'} + T u_t\left(\mathbf{r}\right),
\end{equation}
where $\lambda = 1 - T \alpha$.

@@

\section{Reduction to Finite State-Space Model}\label{Simplifying Decomposition}
Starting from Eq.~\ref{reduced continuous model} we multiply throughout by $\boldsymbol{\phi}(r)$ and integrate over the spatial domain $\Omega$ to get
\begin{eqnarray}\label{Poo}
	\int_\Omega  {\boldsymbol{\phi} \left(\mathbf{r}\right)\boldsymbol{\phi}^{\top}\left(\mathbf{r}\right) d\mathbf{r}} \mathbf{x}_{t+1} = T \int_\Omega  {\boldsymbol{\phi} (\mathbf{r}) \boldsymbol{\theta}^{\top} \int_\Omega  {\boldsymbol{\psi}  \left(\mathbf{r}-\mathbf{r}'\right) f\left(\boldsymbol{\phi}^{\top}\left(\mathbf{r}'\right) \mathbf{x}_t \right)d\mathbf{r}'}d\mathbf{r}} - \lambda\int_\Omega {\boldsymbol{\phi}(\mathbf{r})\boldsymbol{\phi}^{\top}(\mathbf{r})d\mathbf{r}} \mathbf{x}_t \\
	+ T \int_\Omega{\boldsymbol{\phi} \left(\mathbf{r}\right) u_t\left(\mathbf{r}\right)d\mathbf{r}} + \int_\Omega{\boldsymbol{\phi} \left(\mathbf{r}\right) e_t\left(\mathbf{r}\right)d\mathbf{r}}.
\end{eqnarray}
Substituting Eq.~\ref{DefGamma} into Eq.~\ref{Poo} and cross-multiplying by $\boldsymbol{\Gamma}^{-1}$ gives 
\begin{eqnarray}\label{Homogeneous SS Model}
	\mathbf{x}_{t+1} = T\boldsymbol{\Gamma}^{ - 1}\int_\Omega {\boldsymbol{\phi}\left(\mathbf{r}\right) \int_\Omega {f\left(\boldsymbol{\phi}^{\top}\left(\mathbf{r}'\right)\mathbf{x}_t\right) \boldsymbol{\psi}^{\top} \left(\mathbf{r}-\mathbf{r}'\right)d\mathbf{r}'} d\mathbf{r}} \boldsymbol{\theta} - \lambda \mathbf{x}_t \\
	+ \boldsymbol{\Gamma}^{-1}T \int_\Omega{\boldsymbol{\phi} \left(\mathbf{r}\right) u_t\left(\mathbf{r}\right)d\mathbf{r}} + \boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}\left(\mathbf{r}\right)e_t\left(\mathbf{r}\right)d\mathbf{r}}.
\end{eqnarray}

\section{Colored noise}
We have
\begin{equation}
v_{t + 1}\left( \mathbf{r} \right)  = T\int_\Omega  {k\left( {\mathbf{r}} , {\mathbf{r}'} \right)f\left( {{v_t}\left( \mathbf{r'} \right)} \right)d{\mathbf{r'}}} + \lambda v_t\left( {\mathbf{r}} \right) +e_t\left( {\mathbf{r}} \right).
\end{equation}
where the disturbance $e_t(\mathbf{r})$ is a zero-mean normally distributed noise process, spatially coloured but temporally white, with covariance

\begin{equation}
cov(e_t(\mathbf{r}),e_{t+\tau}(\mathbf{r'}))=
\begin{cases}
\beta\left(\mathbf{r}-\mathbf{r'}\right), & \tau=0 \\
0 & \mathrm{otherwise}
\end{cases}
\label{eq:FieldCovariance}
\end{equation}
The final state-space model is given by
\begin{equation}
\mathbf{x}_{t + 1} = T q(\mathbf{x}_t)\boldsymbol{\theta} -\lambda \mathbf{x}_t + \boldsymbol{\Gamma}^{-1}\int_\Omega  {\boldsymbol{\phi} ( \mathbf{r} )e_t( \mathbf{r} )d\mathbf{r}}
\end{equation}
then the covariance of the noise term is given by
\begin{eqnarray}
\mathbf{\Sigma}_w&=&\mathbf{\Gamma}^{-1}\mathbf E[\int_{\Omega}\boldsymbol{\phi}\left(\mathbf{r}\right)e_t\left(\mathbf{r}\right)d\mathbf{r} \int_{\Omega}\boldsymbol{\phi}\left(\mathbf{r}'\right)^{\top} e_t\left(\mathbf{r}'\right)d\mathbf{r}']\mathbf{\Gamma}^{- \top} \nonumber \\
&=&\mathbf{\Gamma}^{-1}\iint\limits_{\Omega} \boldsymbol{\phi}\left(\mathbf{r}\right) \mathbf E[e_t\left(\mathbf{r}\right)e_t\left(\mathbf{r}'\right)]\boldsymbol{\phi}\left(\mathbf{r}'\right)^{\top}d\mathbf{r}' d\mathbf r\mathbf{\Gamma}^{- \top} \nonumber\\
&=&\mathbf{\Gamma}^{-1}\iint\limits_{\Omega}\boldsymbol{\phi}\left(\mathbf r\right) \beta\left(\mathbf r- \mathbf r' \right)\boldsymbol{\phi}\left(\mathbf r'\right)^{\top}d\mathbf r' d\mathbf r\mathbf{\Gamma}^{- \top} 
\end{eqnarray}

\section{Parameter estimation using LS}
\begin{equation}
 \mathbf x_{t+1}=\mathbf{x}_{t}+T_s \mathbf{Q}(\mathbf{x}_t)\boldsymbol{\theta}-\alpha T_s\mathbf{x}_t+T_s\Gamma^{-1}\int_\Omega\boldsymbol{\phi}(\mathbf r)\varepsilon_t(\mathbf r)d\mathbf r
\label{eq:statespacemodel}
\end{equation}
where the matrix $\mathbf Q_{n_x \times n_{\boldsymbol{\theta}}}(\mathbf x_t)$ is defined as
\begin{equation}
 Q_{n_x \times n_{\boldsymbol{\theta}}}=\Gamma^{-1}\int_\Omega\boldsymbol{\phi}(\mathbf r)\int_\Omega f(\boldsymbol{\phi}^\top(\mathbf r')\mathbf x_t)\mathbf{\psi}^\top(\mathbf r-\mathbf r')d\mathbf r'd\mathbf r
\end{equation}
and $T_s$ is the sampling period. Least Square method can be used as (\ref{eq:statespacemodel}) is linear in parameters. At each time instant we have

>>>>

where $T$ is the time step, $\lambda = 1-\alpha T$ and is an iid spatially colored, temporally white disturbance term with the covariance 
\begin{equation}\label{eq:FieldCovariance}
\gamma\left(\mathbf{r}-\mathbf{r'}\right), & \tau=0 \\
 This term was added to drive the system and account for uncertainty in the model. To generate iEEG/LFP data we used the output function 

@@

where $\mathbf{r}_n$ defines the location of the sensors in the field where $n=1,...,N$ indexes the sensors and $\varepsilon_t \sim \mathcal{N}\left(0,\Sigma{\varepsilon}\right)$. The output kernel $m(\mathbf{r}-\mathbf{r}')$ governs the geometry of pick-up range of sensors where

@@

where $\mathbf{e}_t$ is a zero-mean normal white noise term with the covariance (see Appendix \ref{ColoredNoise})

@@

\boldsymbol{\Sigma}_e=\boldsymbol{\Gamma}^{-1}\iint\limits_{\Omega}\boldsymbol{\phi}\left(\mathbf r\right) \rho\left(\mathbf{r}- \mathbf{r}' \right)\boldsymbol{\phi}\left(\mathbf{r}'\right)^{\top}d\mathbf{r}' d\mathbf{r}\boldsymbol{\Gamma}^{- \top} 

@@

The approximation of the state density is made by propagating the so called sigma points through the state-space model. The sigma points describe the mean and covariance of the state estimate. The accuracy of the approximation is linked to the time step used in the temporal discretization of the system, where the state transition appears more linear with a finer resolution. To explain the operation of the UKF we begin by deriving the traditional Kalman filter with the state-space model
\begin{eqnarray}
	\mathbf{x}_t = q(\mathbf{x}_t)\boldsymbol{\theta} - \lambda\mathbf{x}_t + k\left(u_t\left(\mathbf{r}\right)\right) + \mathbf{e}_t \\
	\mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t,
\end{eqnarray}
where $\boldsymbol{\theta}$ and $\lambda$ are the systems parameters. In forwards iteration since the observation equation is linear in parameters we only need to use Unsented transformation in prediction step. The update step equations remains the same as the standard Kalman filtering formulations. In backward smoothing path, estimation from the forward filtering is used to form the new set of sigma points in a similar way to the filterin step, however the predicted cross-covariance of the states need to be calculated. A summary of the algorithm is given in Algorithm 
\begin{algorithm}
\caption{The Unscented Kalman Filter}\label{CPTEMDAlgorithm}
\begin{algorithmic}[1]
\State Initialise
\begin{equation*}
\hat{\mathbf x}_0=\mathbb E[\mathbf x_0]
\end{equation*}
\begin{equation*}
\mathbf P_0=\mathbb E[(\mathbf x_0-\hat{\mathbf x}_0)(\mathbf x_0-\hat{\mathbf x}_0)^\top]
\end{equation*}
\State Form the sigma vectors 
\begin{equation*}
\mathcal X_k=[\hat{\mathbf x}_k \quad \hat{\mathbf x}_k+\gamma\sqrt{\mathbf P_k} \quad \hat{\mathbf x}_k-\gamma\sqrt{\mathbf P_k}]
\end{equation*}
\State Update the sigma vectors by propogating them through state transition equation and compute the predicted state and the predicted state covariance
\begin{equation*}
 \hat{\mathbf x}_{k +1|k}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,k+1|k}
\end{equation*}
\begin{equation*}
 \mathbf P_{k +1|k}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,k+1|k}-\hat{\mathbf x}_{k +1|k})(\mathcal X_{i,k+1|k}-\hat{\mathbf x}_{k +1|k})^\top+\boldsymbol \Sigma_w
\end{equation*}
\State use extrema to form the splines $s_{min,k}(n)$ and $s_{max,k}(n)$
\State calculate $x_{0,k}(n) = 0.5\left(s_{min,k}(n) + s_{max,k}(n)\right)$
\State find local maxima of $h_k(n)$
\State use maxima to form the spline $r_k(n) = s_{max,k}(n)$
\While { $x_{0,k}(n)$ is not monotonic }
    \State $k=k+1$
    \State perform CPT on $x_{0,k-1}(n)$ to estimate $\phi_{k}(n)$
    \State repeats steps 3 to 7 to approximate parameters
\EndWhile

\end{algorithmic}
\end{algorithm}
\section{Comparison of Estimated Parameters and True Parameters}

@@

% \section{Discrete Time Model}\label{Time Discretization}
 \section{Discrete Time Model}\label{Time Discretization}
To form the IDE neural field model a time discretization must be perform. We used a one-step Euler method where Eq.~\ref{FinalForm1} can be approximated by
\begin{equation}\label{Euler Approximation}	
\frac{v\left( \mathbf{r},t+T \right) - v\left( \mathbf{r},t\right)}{T} =   -\alpha v\left( \mathbf{r},t \right) + \int_\Omega  {w\left( \mathbf{r}-\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)d\mathbf{r}'} + u\left(\mathbf{r},t\right).
\end{equation}
For clarity, we shall index time points in the discrete time form of the model using the subscript $t$ and the next time point as $t+1$. Rearranging Eq.~\ref{Euler Approximation} we get
\begin{eqnarray}\label{Euler Approximation}	
v_{t+1}\left( \mathbf{r}\right) &=& v_t\left( \mathbf{r}\right) -T \alpha v_t\left( \mathbf{r}\right) + T \int_\Omega  {w\left( \mathbf{r}-\mathbf{r}' \right)f\left( {v_t\left( \mathbf{r}'\right)} \right)d\mathbf{r}'} + T u_t\left(\mathbf{r}\right).
\end{eqnarray}
The discrete time form of the model is
\begin{equation}\label{Discrete Time Model1}
	v_{t+1}\left(\mathbf{r}\right) = \lambda v_t\left(\mathbf{r}\right) + T \int_\Omega { w\left(\mathbf{r}-\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'} + T u_t\left(\mathbf{r}\right),
\end{equation}
where $\lambda = 1 - T \alpha$.

@@

\section{Reduction to Finite State-Space Model}\label{Simplifying Decomposition}
Starting from Eq.~\ref{reduced continuous model} we multiply throughout by $\boldsymbol{\phi}(r)$ and integrate over the spatial domain $\Omega$ to get
\begin{eqnarray}\label{Poo}
	\int_\Omega  {\boldsymbol{\phi} \left(\mathbf{r}\right)\boldsymbol{\phi}^{\top}\left(\mathbf{r}\right) d\mathbf{r}} \mathbf{x}_{t+1} = T \int_\Omega  {\boldsymbol{\phi} (\mathbf{r}) \boldsymbol{\theta}^{\top} \int_\Omega  {\boldsymbol{\psi}  \left(\mathbf{r}-\mathbf{r}'\right) f\left(\boldsymbol{\phi}^{\top}\left(\mathbf{r}'\right) \mathbf{x}_t \right)d\mathbf{r}'}d\mathbf{r}} - \lambda\int_\Omega {\boldsymbol{\phi}(\mathbf{r})\boldsymbol{\phi}^{\top}(\mathbf{r})d\mathbf{r}} \mathbf{x}_t \\
	+ T \int_\Omega{\boldsymbol{\phi} \left(\mathbf{r}\right) u_t\left(\mathbf{r}\right)d\mathbf{r}} + \int_\Omega{\boldsymbol{\phi} \left(\mathbf{r}\right) e_t\left(\mathbf{r}\right)d\mathbf{r}}.
Substituting Eq.~\ref{DefGamma} into Eq.~\ref{Poo} and cross-multiplying by $\boldsymbol{\Gamma}^{-1}$ gives 
	\mathbf{x}_{t+1} = T\boldsymbol{\Gamma}^{ - 1}\int_\Omega {\boldsymbol{\phi}\left(\mathbf{r}\right) \int_\Omega {f\left(\boldsymbol{\phi}^{\top}\left(\mathbf{r}'\right)\mathbf{x}_t\right) \boldsymbol{\psi}^{\top} \left(\mathbf{r}-\mathbf{r}'\right)d\mathbf{r}'} d\mathbf{r}} \boldsymbol{\theta} - \lambda \mathbf{x}_t \\
	+ \boldsymbol{\Gamma}^{-1}T \int_\Omega{\boldsymbol{\phi} \left(\mathbf{r}\right) u_t\left(\mathbf{r}\right)d\mathbf{r}} + \boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}\left(\mathbf{r}\right)e_t\left(\mathbf{r}\right)d\mathbf{r}}.
\section{}\label{ColoredNoise}
\newtheorem{lemma}{Lemma}
\begin{lemma}
Consider the zero-mean process $e_t\left(\mathbf r\right)$ with covariance as defined by (\ref{eq:FieldCovariance}) then
 \mathbf e_t=\boldsymbol{\Gamma}^{-1}\int_\Omega  {\boldsymbol{\phi} ( \mathbf{r} )e_t( \mathbf{r} )d\mathbf{r}}
\label{eq:AppendixWt}
is a vector valued, zero mean normally distributed white noise process with covariance
\boldsymbol\Sigma_e =\mathbf{\Gamma}^{-1}\iint\limits_{\Omega}\boldsymbol{\phi}\left(\mathbf r\right) \beta\left(\mathbf r- \mathbf r' \right)\boldsymbol{\phi}\left(\mathbf r'\right)^{\top}d\mathbf r' d\mathbf r\mathbf{\Gamma}^{- \top} 
\label{lemma:FieldCovariance}
\end{lemma}
\begin{IEEEproof}
 Equation (\ref{eq:AppendixWt}) is a linear function of $e_t(\mathbf r)$ and hence $\mathbf{e}_t$ is also normally distributed. The expected value of $\mathbf e_t$ is given by 
\begin{eqnarray}
 \mathbf E\left[ \mathbf e_t\right]&=& \mathbf{\Gamma}^{-1}\int_{\Omega}\boldsymbol\phi\left(\mathbf{r}\right)\mathbf E\left[e_t\left(\mathbf{r}\right)\right] d\mathbf{r} \nonumber \\
&=&\mathbf 0
The covariance of $\mathbf{e}_t$ is
\begin{eqnarray}
\mathbf{\Sigma}_e&=&\mathbf{\Gamma}^{-1}\mathbf E[\int_{\Omega}\boldsymbol{\phi}\left(\mathbf{r}\right)e_t\left(\mathbf{r}\right)d\mathbf{r} \int_{\Omega}\boldsymbol{\phi}\left(\mathbf{r}'\right)^{\top} e_t\left(\mathbf{r}'\right)d\mathbf{r}']\mathbf{\Gamma}^{- \top} \nonumber \\
&=&\mathbf{\Gamma}^{-1}\iint\limits_{\Omega} \boldsymbol{\phi}\left(\mathbf{r}\right) \mathbf E[e_t\left(\mathbf{r}\right)e_t\left(\mathbf{r}'\right)]\boldsymbol{\phi}\left(\mathbf{r}'\right)^{\top}d\mathbf{r}' d\mathbf r\mathbf{\Gamma}^{- \top} \nonumber\\
&=&\mathbf{\Gamma}^{-1}\iint\limits_{\Omega}\boldsymbol{\phi}\left(\mathbf r\right) \gamma\left(\mathbf r- \mathbf r' \right)\boldsymbol{\phi}\left(\mathbf r'\right)^{\top}d\mathbf r' d\mathbf r\mathbf{\Gamma}^{- \top} 
\end{IEEEproof}
\section{Parameter estimation using LS}
 \mathbf x_{t+1}=\mathbf{x}_{t}+T_s \mathbf{Q}(\mathbf{x}_t)\boldsymbol{\theta}-\alpha T_s\mathbf{x}_t+T_s\Gamma^{-1}\int_\Omega\boldsymbol{\phi}(\mathbf r)\varepsilon_t(\mathbf r)d\mathbf r
\label{eq:statespacemodel}
where the matrix $\mathbf Q_{n_x \times n_{\boldsymbol{\theta}}}(\mathbf x_t)$ is defined as
 Q_{n_x \times n_{\boldsymbol{\theta}}}=\Gamma^{-1}\int_\Omega\boldsymbol{\phi}(\mathbf r)\int_\Omega f(\boldsymbol{\phi}^\top(\mathbf r')\mathbf x_t)\mathbf{\psi}^\top(\mathbf r-\mathbf r')d\mathbf r'd\mathbf r
and $T_s$ is the sampling period. Least Square method can be used as (\ref{eq:statespacemodel}) is linear in parameters. At each time instant we have
